---
title: "Gasoline prices in Japan"
author: "Mitsuo Shiota"
date: "2021-11-17"
output: 
  github_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(readxl)
library(tsibble)
library(fable)
library(lubridate)
library(pdftools)
library(patchwork)
library(tidyquant)
```

Updated: `r Sys.Date()`

## Summary

[Nikkei reported](https://www.nikkei.com/article/DGXZQOUA169350W1A111C2000000/) on November 17, 2021 that Ministry of Economy, Trade and Industry (METI) is considering a subsidy to gasoline wholesalers to suppress gasoline retail prices when they rise to higher than 170 yen per litre. At the end of the article, Nikkei showed a skeptical view saying, "As the number of retailers has decreased, the retailers may not reduce retail prices even if wholesale prices fall." So I study the relationship among retail and wholesale prices of gasoline and imported crude oil prices in Japan.

I find the current surge in retail prices of gasoline is mainly due to the surge in imported crude oil prices, and partly due to the reduced competition among retailers and wholesalers, who are getting more margins.

To give a subsidy to wholesalers may incentivize them to raise their margins even more. This subsidy idea is contrary to that of the Biden Administration, which has begun to investigate oil companies.

## Get gasoline prices

Agency for National Resources Energy under METI publishes gasoline prices in its [web site](https://www.enecho.meti.go.jp/statistics/petroleum_and_lpgas/pl007/results.html#headline1). Although the original retail price data include consumption tax since April 1, 2004, I exclude consumption tax all over the period.

```{r get_data, message=FALSE}

wholesale <- read_excel("data/220228o5.xls",
                        sheet = "レギュラー",
                        col_types = c("text", "date", rep("text", 55))
                        ) %>% 
  select(date = "調査年月", price = "全         国") %>% 
  mutate(
    date = as.Date(date),
    price = as.numeric(price)
    ) 

wholesale_price <- wholesale %>% 
  mutate(month = yearmonth(date)) %>% 
  select(month, wholesale_price = price) %>% 
  as_tsibble(index = month)

retail <- read_excel("data/220330s5.xls", 
                     sheet = "レギュラー",
                        col_types = c("text", "date", rep("text", 59)))

retail <- retail %>% 
  slice_head(n = nrow(retail) - 1) %>% 
  select(date = "調査日", price = "全         国") %>% 
  mutate(
    date = as.Date(date),
    price = as.numeric(price)
    ) 

retail_price <- retail %>% 
  mutate(
    tax_rate = case_when(
      date < "2004-04-01" ~ 0,
      date < "2014-04-01" ~ .05,
      date < "2019-10-01" ~ .08,
      TRUE ~ .1
                       ),
    price_excl_tax = price / (1 + tax_rate)
  )

retail_price_month <- retail_price %>% 
  select(date, retail_price = price_excl_tax) %>% 
  as_tsibble(index = date) %>% 
  index_by(month = yearmonth(date)) %>% 
  summarize(retail_price = mean(retail_price))

```

I get imported crude oil price data from customs statistics via [e-Stat](https://www.e-stat.go.jp/stat-search/files?page=1&layout=datalist&toukei=00350300&bunya_l=16&tstat=000001013141&cycle=1&tclass1=000001013192&tclass2=000001013194&tclass3val=0). 

```{r get_crude_from_csv, message=FALSE}

get_crude_oil_price <- function(df) {
  df %>% 
    mutate(Commodity =  str_remove_all(Commodity, "'| ")) %>% 
    filter(Commodity == "30301") %>% 
    pivot_longer(starts_with("Quantity-"), names_to = "month", values_to = "quantity") %>% 
    pivot_longer(starts_with("Value-")) %>% 
    mutate(
      month = str_remove(month, "Quantity-"),
      name = str_remove(name, "Value-")
    ) %>% 
    filter(month == name, month != "Year") %>% 
    mutate(
      month = if_else(month == "Apl", "Apr", month),
      date = dmy(paste(1, month, Year)),
      crude_oil_price = value / quantity
    ) %>% 
    select(date, crude_oil_price)
}

import <- paste0("data/", dir("data/", ".csv")) %>% 
  map(~ read_csv(.x) %>% get_crude_oil_price) %>% 
  bind_rows() %>% 
  filter(!is.na(crude_oil_price))

```

```{r get_crude_from_pdf, eval=FALSE}

get_from_pdf <- function(id) {
  paste0("https://www.e-stat.go.jp/stat-search/file-download?statInfId=00000", id, "&fileKind=2") %>% 
    pdf_text() %>% 
    read_csv() %>% 
    setNames(c("line")) %>% 
    filter(str_detect(line, "^\\d")) %>% 
    pull(line) %>% 
    paste(collapse = "\n") %>% 
    read_fwf(col_positions = fwf_widths(c(23, 2, 15, 17, 17, 11)),
             col_types = cols(.default = col_character())) %>% 
    mutate(statInfId = id)
}

id_date <- tibble(
  statInfId = c("2356861", "2360681", "2366821", # 2001
               "2370903", "2374447", "2377215",
               "2388625", "2395224", "2400500",
               "2403671", "2409041", "2413944",
               "2353533", "2368407", "2373904", # 2002
               "2390257", "2397348", "2406335",
               "2414755", "2419894", "2434540",
               "2462003", "2477263", "2487568",
               "2154280", "2159214", "2167204", # 2003
               "2172358", "2182731", "2187465",
               "2192337", "2196921", "2203979",
               "2206662", "2210342", "2214409",
               "2154829", "2163290", "2170020", # 2004
               "2176027", "2184517", "2193631",
               "2212517", "2216694", "2224086",
               "2243822", "2253031", "2263407",
               "2221012", "2225946", "2228322", # 2005
               "2231036", "2234710", "2238784",
               "2243156", "2248685", "2251982",
               "2254626", "2257154", "2263286",
               "2070609", "2073232", "2075826", # 2006
               "2078138", "2080876", "2084239",
               "2086779", "2089357", "2091820",
               "2094179", "2096541", "2099394",
               "2101765", "2104568", "2107778", # 2007
               "2110529", "2113720", "2116186",
               "2119134", "2122199", "2125080",
               "2127655", "2130968", "2133192",
               "2524275", "2524925", "2525183", # 2008
               "2525808", "2526624", "2527557",
               "2528619", "2530117", "2530804", 
               "2534412", "2531505", "2532132"
               ),
  date = map(2001:2008, ~ make_date(year = .x, month = 1:12)) %>% 
      unlist() %>% 
      as.Date(origin = "1970-01-01")
)

import_pdf <- map(id_date$statInfId, ~ get_from_pdf(.x)) %>% 
  bind_rows() %>% 
  left_join(id_date, by = "statInfId")

save(import_pdf, file = "data/import_from_pdf.rda")

```

```{r load_pdf_data}

load(file = "data/import_from_pdf.rda")

```

```{r combine}

import_price <- import_pdf %>% 
  filter(X1 == "30301") %>% 
  select(date, quantity = X3, value = X4) %>% 
  mutate(
    across(quantity:value, as.numeric),
    crude_oil_price = value / quantity
    ) %>% 
  select(date, crude_oil_price) %>% 
  bind_rows(import)  %>% 
  mutate(month = yearmonth(date)) %>% 
  as_tsibble(index = month) %>% 
  select(-date)

combo <- retail_price_month %>% 
  filter_index("2000 Jul" ~ .) %>% 
  left_join(wholesale_price, by = "month") %>% 
  left_join(import_price, by = "month") %>% 
  mutate(
    crude_plus_tax = crude_oil_price + 53.8,
    diff_rw = retail_price - wholesale_price,
    diff_wc = wholesale_price - crude_plus_tax
    ) %>% 
  pivot_longer(retail_price:diff_wc) %>% 
  update_tsibble(key = name)

```

## Plot retail, wholesale gas and crude oil prices

Yes, both retail and wholesale prices are rising, mainly due to rising imported crude oil prices.

```{r plot1, warning=FALSE}

combo %>% 
  filter(name %in% c("retail_price", "wholesale_price", "crude_plus_tax")) %>% 
  autoplot(value) +
  geom_hline(yintercept = 170/1.1, color = "gray50") +
  annotate("text", x = yearmonth("2010 Jan"), y = 160, label = "155 (170 including consumption tax)", size = 4) +
  geom_hline(yintercept = 53.8, color = "gray50") +
  annotate("text", x = yearmonth("2010 Jan"), y = 59, label = "53.8 gasoline tax", size = 4) +
  annotate("text", x = yearmonth("2000 Jan"), y = 110, label = "Retail", size = 4) +
  annotate("text", x = yearmonth("2001 Jan"), y = 90, label = "Wholesale", size = 4) +
  annotate("text", x = yearmonth("2003 Jan"), y = 70, label = "Crude oil (right axis)", size = 4) +expand_limits(y = 0) +
  scale_y_continuous(
    name = "Yen per litre\n(excluding consumption tax)",
    sec.axis = sec_axis(~ .x - 53.8)
  ) +
  labs(title = "Gasoline and crude oil prices in Japan") +
  theme(legend.position = "none")

```

## Plot price differences

Differences between retail and wholesale prices have been increasing since 2016. The average difference was `r combo %>% filter(name == "diff_rw") %>% filter_index("2000 Jul" ~ "2015 Dec") %>% pull(value) %>% mean() %>% round()` yen per litre from 2000 Jul to 2015 Dec, and is `r combo %>% filter(name == "diff_rw", !is.na(value)) %>% slice_tail(n = 1) %>% pull(value) %>% round()` in `r combo %>% filter(name == "diff_rw", !is.na(value)) %>% slice_tail(n = 1) %>% pull(month)`. This may reflect the reduced competition among retailers. You can see the number of retailers has constantly decreased since around 1995 in the last page of [this material (Japanese)](https://www.enecho.meti.go.jp/category/resources_and_fuel/distribution/hinnkakuhou/data/2021_07_30_01.pdf) from Agency for National Resources Energy.

```{r plot 2, warning=FALSE}

combo %>% 
  filter(name == "diff_rw") %>% 
  autoplot(value) +
  expand_limits(y = 0) +
  labs(y = "Yen per litre",
       title = "Differences between retail and wholesale prices")

```

Differences between wholesale price and imported crude oil price plus gasoline tax have also been increasing. The average difference was `r combo %>% filter(name == "diff_wc") %>% filter_index("2001 Jan" ~ "2014 Dec") %>% pull(value) %>% mean() %>% round()` yen per litre from 2001 Jan to 2014 Dec, and is `r combo %>% filter(name == "diff_wc", !is.na(value)) %>% slice_tail(n = 1) %>% pull(value) %>% round()` in `r combo %>% filter(name == "diff_wc", !is.na(value)) %>% slice_tail(n = 1) %>% pull(month)`. This may reflect the reduced competition among wholesalers, who have got monopolistic power by consolidation.

```{r plot3, warning=FALSE}

combo %>% 
  filter(name == "diff_wc") %>% 
  autoplot(value) +
  labs(y = "Yen per litre",
       title = "Differences between wholesale price and\nimported crude oil price + gasoline tax")

```

## Dubai crude oil price (monthly)

Import prices pretty precisely follow Dubai crude oil spot prices of one month ago.

```{r get_dubai_data}

dub_oil <- tq_get(c("POILDUBUSDM", "EXJPUS"), get = "economic.data", from = "1990-01-01")

dub_oil_month <- dub_oil %>% 
  pivot_wider(names_from = symbol, values_from = price) %>% 
  mutate(dub_oil_price = POILDUBUSDM * EXJPUS / 158.99) %>% 
  select(date, dub_oil_price) %>% 
  mutate(month = yearmonth(date)) %>% 
  as_tsibble(index = month) %>% 
  select(-date)
```


```{r}

dub_import <- dub_oil_month %>% 
  inner_join(import_price, by = "month") %>% 
  rename(import_price = crude_oil_price) %>% 
  filter(!is.na(dub_oil_price), !is.na(import_price))

dub_import %>% 
  pivot_longer(c(import_price, dub_oil_price)) %>% 
  autoplot(value)

dub_import %>% 
  mutate(lag_dub = lag(dub_oil_price)) %>% 
  select(-dub_oil_price) %>% 
  pivot_longer(c(import_price, lag_dub)) %>% 
  autoplot(value)

p0 <- dub_import %>% 
  ggplot(aes(dub_oil_price, import_price)) +
  geom_point(alpha = 0.2) +
  coord_fixed()

p1 <- dub_import %>% 
  ggplot(aes(lag(dub_oil_price, 1), import_price)) +
  geom_point(alpha = 0.2) +
  coord_fixed()

p2 <- dub_import %>% 
  ggplot(aes(lag(dub_oil_price, 2), import_price)) +
  geom_point(alpha = 0.2) +
  coord_fixed()

p0 | p1 | p2

```

Correlations are `r cor(dub_import$dub_oil_price, dub_import$import_price)` if 0 month lag from Dubai to Japan customs, `r cor(dub_import$dub_oil_price[-251], dub_import$import_price[-1])` if 1 month lag, and `r cor(dub_import$dub_oil_price[c(-250, -251)], dub_import$import_price[c(-1, -2)])` if 2 months lag.

## Dubai crude oil price (weekly)

I get daily data from [oilprice.com](https://oilprice.com/jp/%E5%8E%9F%E6%B2%B9%E4%BE%A1%E6%A0%BC%E3%83%81%E3%83%A3%E3%83%BC%E3%83%88).

```{r get_daily_dubai}

dubai_daily <- tribble(
  ~date, ~price,
  "2021-11-22", 80.548,
  "2021-11-24", 81.201,
  "2021-11-26", 80.084,
  "2021-11-29", 80.305,
  "2021-11-30", 80.276,
  "2021-12-02", 80.276,
  "2021-12-03", 68.909,
  "2021-12-06", 71.654,
  "2021-12-08", 73.693,
  "2021-12-10", 73.189,
  "2021-12-14", 72.223,
  "2021-12-15", 72.383,
  "2021-12-17", 72.432,
  "2021-12-20", 71.655,
  "2021-12-22", 72.71,
  "2021-12-27", 73.2,
  "2021-12-30", 73.19,
  "2021-12-31", 73.185,
  "2022-01-03", 73.185,
  "2022-01-04", 73.185,
  "2022-01-05", 78.812,
  "2022-01-06", 79.695,
  "2022-01-07", 79.535,
  "2022-01-07", 79.535,
  "2022-01-10", 78.887,
  "2022-01-11", 80.789,
  "2022-01-12", 81.342,
  "2022-01-13", 81.297,
  "2022-01-14", 82.269,
  "2022-01-18", 83.221,
  "2022-01-19", 83.559,
  "2022-01-20", 83.496,
  "2022-01-21", 83.291,
  "2022-01-24", 82.868,
  "2022-01-25", 83.213,
  "2022-01-26", 83.389,
  "2022-01-27", 83.333,
  "2022-01-28", 83.362,
  "2022-01-31", 83.455,
  "2022-02-01", 83.455,
  "2022-02-04", 91.056,
  "2022-02-07", 90.757,
  "2022-02-08", 89.282,
  "2022-02-09", 89.883,
  "2022-02-10", 89.918,
  "2022-02-11", 91.79,
  "2022-02-14", 92.605,
  "2022-02-15", 90.859,
  "2022-02-16", 91.558,
  "2022-02-18", 90.897,
  "2022-02-22", 91.838,
  "2022-02-24", 92.074,
  "2022-02-25", 91.928,
  "2022-02-28", 92.34,
  "2022-03-01", 92.34,
  "2022-03-02", 92.34,
  "2022-03-03", 106.578,
  "2022-03-04", 114.95,
  "2022-03-07", 118.18,
  "2022-03-08", 122.532,
  "2022-03-09", 110.86
) %>% 
  mutate(date = as.Date(date))

yen_dollar_daily <- tq_get(c("DEXJPUS"), get = "economic.data", from = "2020-01-01") %>% 
  select(-symbol)

yen_dollar_daily <- yen_dollar_daily %>% 
  rename(yen_dollar_rate = price)

dubai_daily2 <- dubai_daily %>% 
  left_join(yen_dollar_daily, by = "date") %>% 
  mutate(dubai_oil_price = price * yen_dollar_rate / 158.99) %>% 
  select(date, dubai_oil_price)

dubai_weekly <- dubai_daily2 %>% 
  mutate(week = yearweek(date)) %>% 
  group_by(week) %>% 
  summarize(price = mean(dubai_oil_price, na.rm = TRUE))

```

[METI publishes](https://nenryo-gekihenkanwa.jp/pdf/result_rev8.pdf) Dubai crude oil prices, counterfactual gasoline prices without subsidies, actual gasoline prieces and effects of subsidies, which are differences between counterfacual and actual, all in the units of yen / litre every week.

```{r meti_calc}

meti_dubai_weekly <- tribble(
  ~start_date, ~price,
  "2022-01-03", 57.8,
  "2022-01-10", 59.4,
  "2022-01-17", 62.6,
  "2022-01-24", 62.9,
  "2022-01-31", 64.4,
  "2022-02-07", 66.2,
  "2022-02-14", 67.4,
  "2022-02-22", 69.9,
  "2022-03-01", 81.6,
  "2022-03-08", 85.8,
  "2022-03-15", 76.8,
  "2022-03-22", 88
) %>% 
  mutate(week = yearweek(start_date)) %>% 
  select(-start_date)

meti_subsidy <- tribble(
  ~start_date, ~subsidy,
  "2022-01-17", 3.4,
  "2022-01-24", 3.7,
  "2022-01-31", 5,
  "2022-02-07", 5,
  "2022-02-14", 5,
  "2022-02-22", 5,
  "2022-03-01", 17.7,
  "2022-03-08", 25,
  "2022-03-15", 18.6,
  "2022-03-22", 25
) %>% 
  mutate(week = yearweek(start_date)) %>% 
  select(-start_date)
```

Weekly prices are not so different between oilprice.com and METI.

```{r dubai_chart}

dubai_weekly %>% 
  mutate(source = "oilprice.com") %>% 
  bind_rows(meti_dubai_weekly %>% mutate(source = "METI")) %>% 
  ggplot(aes(week, price, color = source)) +
  geom_line() +
  scale_x_yearweek(date_labels = "%b %d") +
  labs(x = NULL, y = "Yen per liter", color = "Data source",
       title = "Dubai crude oil prices (weekly)")
```

So I create weekly prices data by combining oilprice.com in 2021 and METI in 2022.

```{r dubai_weekly}

dubai_weekly2 <- dubai_weekly %>% 
  filter(year(week) == 2021) %>% 
  bind_rows(meti_dubai_weekly %>% filter(year(week) == 2022)) %>% 
  rename(dubai = price)

retail_price_weekly <- retail_price %>% 
  mutate(week = yearweek(date)) %>% 
  select(week, actual = price) %>% 
  left_join(dubai_weekly2, by = "week") %>% 
  filter(!is.na(dubai)) %>% 
  left_join(meti_subsidy, by = "week")

effect_by_lag <- function(lag) {
  retail_price_weekly %>% 
    mutate(
      dubai_lag_n = lag(dubai, lag),
      diff_retail_crude = actual[week == yearweek("2022 W04")] - 
        dubai_lag_n[week == yearweek("2022 W04")],
      counterfactual_wo_subsidy = dubai_lag_n + diff_retail_crude,
      effect_by_subsidy = counterfactual_wo_subsidy - actual
    ) %>% 
    mutate(counterfactual_with_subsidy = counterfactual_wo_subsidy - subsidy) %>% 
    filter(week >= yearweek("2022 W04"))
}

effects <- map_dfr(0:4, effect_by_lag, .id = "lag") %>% 
  mutate(
    lag = as.numeric(lag),
    lag = lag - 1
  )

```

METI calculates subsidy effects on retail prices by assuming 2 weeks lag between Dubai crude oil prices and retail gasoline prices. If we assume 3 weeks lag instead, subsidy effects are smaller. In either case, effects are smaller than subsidies. It means that subsidies are not full reflected in retail prices, and that wholesalers and/or retailers are getting benefits from subsidies.

```{r actual_counterfactual}

effects %>% 
  pivot_longer(c(actual, counterfactual_with_subsidy, counterfactual_wo_subsidy)) %>% 
  mutate(
    name = recode(name,
                  actual = "Actual prices",
                  counterfactual_wo_subsidy = "Counterfactual prices\nif no subsidy",
                  counterfactual_with_subsidy = "Prices if subsidies are\nfully reflected")
  ) %>% 
  ggplot(aes(week, value, color = name)) +
  geom_line() +
  scale_x_yearweek(date_labels = "%b", date_breaks = "1 month") +
  facet_wrap(vars(lag)) +
  labs(x = NULL, y = "Yen per litre", color = NULL,
       title = "Actual and counterfactual retail prices\nby # of week lags from Dubai to retail")

```


```{r subsidy_effect}

effects %>% 
  pivot_longer(c(subsidy, effect_by_subsidy)) %>% 
  mutate(
    name = recode(name,
                  effect_by_subsidy = "Effect",
                  subsidy = "Subsidy")
  ) %>% 
  ggplot(aes(week, value, fill = name)) +
  geom_col(position = "identity", alpha = 0.5) +
  scale_fill_manual(values = c("red", "blue")) +
  scale_x_yearweek(date_labels = "%b", date_breaks = "1 month") +
  facet_wrap(vars(lag)) +
  labs(x = NULL, y = "Yen per litre", fill = NULL,
       title = "Effect of subsidies by # of week lags from Dubai to retail")

```


EOL
